Automated Assessment of Residual Plots with Computer Vision Models

### 2025-07-28 cover_letter/letter.tex

\usepackage{eso-pic} % for background image
\usepackage{graphicx} % necessary for images
\usepackage{xifthen}
\usepackage{parskip} % for spacing
\usepackage{fancyhdr} % for fancy headers
\usepackage{lastpage} 
\usepackage{bera} % for font
\pagestyle{fancy}
\renewcommand{\headrulewidth}{0pt}
\fancyhead{} % footer for pages 2 on
\lfoot{} % Left footer
\cfoot{} % Change for center footer
\rfoot{\footnotesize Page \thepage\ of \pageref{LastPage}} % Right footer page #s 
\renewcommand{\footrulewidth}{0pt}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Letter from ANU},
  pdfauthor={\hspace{0pt}},
  pdfsubject={Submission of Manuscript: Automated Assessment of Residual
Plots with Computer Vision Models},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Letter from ANU}
\author{\hspace{0pt}}
\date{July 25, 2025}
\begin{document}
\AddToShipoutPicture{\ifthenelse{\value{page}<2}{\includegraphics[width=\paperwidth,height=\paperheight]{\_extensions/anuopensci/quarto-anu-letter/assets/images/coverpage.jpg}}{\includegraphics[width=\paperwidth,height=\paperheight]{\_extensions/anuopensci/quarto-anu-letter/assets/images/page-background.jpg}}}

\signature{
\hspace{0pt}}
\address{Australian National University\\Research School of Finance,
Actuarial Studies and Statistics\\26C Kingsley
Street\\Canberra 2600\\Australia\\[2mm]}
\begin{letter}{Editors\\Journal of Computational and Graphical
Statistics\\ ~ \\Subject: Submission of Manuscript: Automated Assessment
of Residual Plots with Computer Vision Models}
\opening{Dear Prof.~Chen and Prof.~Sangalli,}

@@

This paper builds on our previous work, ``A Plot is Worth a Thousand
Tests: Assessing Residual Diagnostics with the Lineup Protocol'' (JCGS),
which provided empirical evidence for the effectiveness of visual
evaluation of residual plots through crowd-sourced lineup experiments.
While that study highlighted the cost and effort of human evaluations,
this work advances automation of residual plot diagnostics through an
application of a computer vision model. We specifically trained a model
to predict a custom distance measure that quantifies the discrepancy
between residuals and theoretically ``good'' distributions. Building on
these predictions, we construct a statistical testing framework aligned
with the original lineup protocol, ensuring a valid visual test. When
evaluated on the earlier experimenal data, the model performs slightly
worse than humans but surpasses conventional statistical tests due to
its reduced sensitivity to minor deviations from model assumptions. We
further demonstrate that the model, like human observers, relies on
visual patterns and shapes, making it a practical tool for reducing
manual effort in residual plot diagnostics. This approach would be of
particular interest to readers concerned with model diagnostics and the
automation of visual data analysis.

>>>>

\usepackage{eso-pic} % for background image
\usepackage{graphicx} % necessary for images
\usepackage{xifthen}
\usepackage{parskip} % for spacing
\usepackage{fancyhdr} % for fancy headers
\usepackage{lastpage} 
\usepackage{bera} % for font
\pagestyle{fancy}
\renewcommand{\headrulewidth}{0pt}
\fancyhead{} % footer for pages 2 on
\lfoot{} % Left footer
\cfoot{} % Change for center footer
\rfoot{\footnotesize Page \thepage\ of \pageref{LastPage}} % Right footer page #s 
\renewcommand{\footrulewidth}{0pt}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Letter from ANU},
  pdfauthor={\hspace{0pt}},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Letter from ANU}
\author{\hspace{0pt}}
\date{July 28, 2025}
\begin{document}
\AddToShipoutPicture{\ifthenelse{\value{page}<2}{\includegraphics[width=\paperwidth,height=\paperheight]{\_extensions/anuopensci/quarto-anu-letter/assets/images/coverpage.jpg}}{\includegraphics[width=\paperwidth,height=\paperheight]{\_extensions/anuopensci/quarto-anu-letter/assets/images/page-background.jpg}}}

\signature{
\hspace{0pt}}
\address{Australian National University\\Research School of Finance,
Actuarial Studies and Statistics\\26C Kingsley
Street\\Canberra 2600\\Australia\\[2mm]}
\begin{letter}{Editors,\\Journal of Computational and Graphical
Statistics}
\opening{Dear Prof.~Chen and Prof.~Sangalli,}

@@

This paper describes a new computer vision model that can automatically
read residual plots, it's architecture and performance relative to human
evaluation and classical hypothesis testing. Our previous work, ``A Plot
is Worth a Thousand Tests: Assessing Residual Diagnostics with the
Lineup Protocol'' (JCGS, 2024), described interesting findings about
human evaluations of residual plots relative to classical hypothesis
testing, discovered when building training samples for the computer
vision model described in our new paper. The computer vision model
described in this new submission was our primary research goal.

We specifically trained a custom model to predict a specified distance
measure that quantifies the discrepancy between residuals and
theoretically ``good'' distributions. Building on these predictions, we
construct a statistical testing framework aligned with the original
lineup protocol, ensuring a valid visual test. Despite substantial
experimentation, we could not manage to get a computer vision model to
perform quite as well as the human evaluations. It is slightly more
sensitive, flagging more residual plots as bad. However, as a computer
model, it can process substantially more residual plots than is possible
through human evaluation. Relative to statistical tests (Breusch-Pagan,
Ramsey RESET, Shapiro-Wilk) it performs substantially better, is less
sensitive to small departures, and jointly assesses a variety of
departures (heteroscedasticity, non-linearity, non-normality).

The paper also shows that the model, like human observers, relies on
visual patterns and shapes, making it a practical tool for reducing
manual effort in residual plot diagnostics. This approach would be of
particular interest to readers concerned with model diagnostics and the
automation of visual data analysis. It represents a new direction for
statistical graphics research that engages with new computational tools.
There is a shiny app and R package available to support access and use
of the computer vision model, and this is documented in a special issue
of the Australian and New Zealand Journal of Statistics that celebrates
25 years of R, to appear later this year.

### 2025-07-28 paper.Rmd

>>>>

The concept of visual inference, as proposed by @buja2009statistical, provides an inferential framework to assess whether residual plots indeed contain visual patterns inconsistent with the model assumptions. The fundamental idea involves testing whether the true residual plot visually differs significantly from null plots, where null plots are plotted with residuals generated from the residual rotation distribution [@langsrud2005rotation], which is a distribution consistent with the null hypothesis $H_0$ that the linear regression model is correctly specified. Typically, the visual test is accomplished through the lineup protocol, where the true residual plot is embedded within a lineup alongside several null plots. If the true residual plot can be distinguished from the lineup, it provides evidence for rejecting $H_0$.

A practical limitation of the lineup protocol lies in its reliance on human judgements [see @li2024plot about the practical limitations]. Unlike conventional statistical tests that can be performed computationally in statistical software, the lineup protocol requires human evaluation of images. This characteristic makes it less suitable for large-scale applications, given the associated high labour costs and time requirements. There is a substantial need to develop an approach to substitute these human judgements with an automated reading of data plots using machines.

Efforts to automate plot interpretation date back to early work like Tukey’s "Scagnostics" [@tukey1985computer], which is a set of numerical statistics that summarize features of scatter plots. @wilkinson2005graph expanded on this work, introducing scagnostics based on computable measures applied to planar proximity graphs. These measures, including, but not limited to, "Outlying", "Skinny", "Stringy", "Straight", "Monotonic", "Skewed", "Clumpy", and "Striated", aimed to characterize outliers, shape, density, trend, coherence and other characteristics of the data. However, as noted by @buja2009statistical, such measures may not capture the full range of visual patterns necessary for residual diagnostics. A more promising direction lies in enabling computers to learn the relevant visual features directly mirroring how humans intuitively interpret plots.

Modern computer vision models are well-suited for addressing this challenge. They rely on deep neural networks with convolutional layers [@fukushima1982neocognitron]. These layers use small, sliding windows to scan the image, performing a dot product to extract local features and patterns. Numerous studies have demonstrated the efficacy of convolutional layers in addressing various vision tasks, including image recognition [@rawat2017deep]. Despite the widespread use of computer vision models in fields like computer-aided diagnosis [@lee2015image], pedestrian detection [@brunetti2018computer], and facial recognition [@emami2012facial], their application in reading data plots remains limited. While some studies have explored the use of computer vision models for tasks such as reading recurrence plots for time series regression [@ojeda2020multivariate], time series classification [@chu2019automatic; @hailesilassie2019financial; @hatami2018classification; @zhang2020encoding], anomaly detection [@chen2020convolutional], and pairwise causality analysis [@singh2017deep], the application of reading residual plots with computer vision models is a new field of study.

### 2025-07-28 paper.tex

@@

Consistent with this, the computer vision model assigns a distance
\(\hat{D}\) above the 95\% quantile of the null distribution
(\(p-\text{value} = 0.00498\)), leading to rejection of \(H_0\). The
bootstrapped distribution further supports this, with most fitted models
also being rejected. In contrast, the RESET and Breusch-Pagan tests
yield \(p\)-values above \(0.3\), suggesting no violation. Only the
Shapiro-Wilk test rejects normality with a small \(p\)-value.

>>>>

The concept of visual inference, as proposed by
\citet{buja2009statistical}, provides an inferential framework to assess
whether residual plots indeed contain visual patterns inconsistent with
the model assumptions. The fundamental idea involves testing whether the
true residual plot visually differs significantly from null plots, where
null plots are plotted with residuals generated from the residual
rotation distribution \citep{langsrud2005rotation}, which is a
distribution consistent with the null hypothesis \(H_0\) that the linear
regression model is correctly specified. Typically, the visual test is
accomplished through the lineup protocol, where the true residual plot
is embedded within a lineup alongside several null plots. If the true
residual plot can be distinguished from the lineup, it provides evidence
for rejecting \(H_0\).

A practical limitation of the lineup protocol lies in its reliance on
human judgements \citep[see][ about the practical
limitations]{li2024plot}. Unlike conventional statistical tests that can
be performed computationally in statistical software, the lineup
protocol requires human evaluation of images. This characteristic makes
it less suitable for large-scale applications, given the associated high
labour costs and time requirements. There is a substantial need to
develop an approach to substitute these human judgements with an
automated reading of data plots using machines.

Efforts to automate plot interpretation date back to early work like
Tukey's ``Scagnostics'' \citep{tukey1985computer}, which is a set of
numerical statistics that summarize features of scatter plots.
\citet{wilkinson2005graph} expanded on this work, introducing
scagnostics based on computable measures applied to planar proximity
graphs. These measures, including, but not limited to, ``Outlying'',
``Skinny'', ``Stringy'', ``Straight'', ``Monotonic'', ``Skewed'',
``Clumpy'', and ``Striated'', aimed to characterize outliers, shape,
density, trend, coherence and other characteristics of the data.
However, as noted by \citet{buja2009statistical}, such measures may not
capture the full range of visual patterns necessary for residual
diagnostics. A more promising direction lies in enabling computers to
learn the relevant visual features directly mirroring how humans
intuitively interpret plots.

Modern computer vision models are well-suited for addressing this
challenge. They rely on deep neural networks with convolutional layers
\citep{fukushima1982neocognitron}. These layers use small, sliding
windows to scan the image, performing a dot product to extract local
features and patterns. Numerous studies have demonstrated the efficacy
of convolutional layers in addressing various vision tasks, including
image recognition \citep{rawat2017deep}. Despite the widespread use of
computer vision models in fields like computer-aided diagnosis
\citep{lee2015image}, pedestrian detection \citep{brunetti2018computer},
and facial recognition \citep{emami2012facial}, their application in
reading data plots remains limited. While some studies have explored the
use of computer vision models for tasks such as reading recurrence plots
for time series regression \citep{ojeda2020multivariate}, time series
classification
\citep{chu2019automatic, hailesilassie2019financial, hatami2018classification, zhang2020encoding},
anomaly detection \citep{chen2020convolutional}, and pairwise causality
analysis \citep{singh2017deep}, the application of reading residual
plots with computer vision models is a new field of study.

@@

Consistent with this, the computer vision model assigns a distance
\(\hat{D}\) above the 95\% quantile of the null distribution
(\(p\text{-value} = 0.00498\)), leading to rejection of \(H_0\). The
bootstrapped distribution further supports this, with most fitted models
also being rejected. In contrast, the RESET and Breusch-Pagan tests
yield \(p\)-values above \(0.3\), suggesting no violation. Only the
Shapiro-Wilk test rejects normality with a small \(p\)-value.
